SU($n$) ferromagnets near the Heisenberg critical point

### 2020-04-30 collective_sun/sun_perturbation.tex
minor cleanup of some text

\let\var\relax
\DeclareMathOperator{\var}{var}
\DeclareMathOperator{\cov}{cov}
\DeclareMathOperator{\diag}{diag}

@@

In this work, we generalize some of the techniques used to harness
SU(2)-symmetric spin interactions for metrological advantage to the
case of SU($n$).  In doing so, we develop a low-energy effective
theory of gapped SU($n$) ferromagnets with weakly broken SU($n$)
symmetry, e.g.~by the presence of external fields or additional
interaction terms.  The restriction of our theory to ferromagnetic
interactions can be relaxed to the assumption of
\begin{enumerate*}
\item an initial state in the manifold of permutationally-symmetric
  states, a multi-level generalization of the Dicke manifold for
  two-level spins, together with the assumption that
\item this manifold is gapped away from states that break
  permutational symmetry.
\end{enumerate*}
In addition to working out the effective Hamiltonian induced by weak
SU($n$) symmetry breaking through second order in perturbation theory,
we develop a diagrammatic method to systematically compute all matrix
elements of arbitrary spin operators restricted manifolds of weakly
broken permutation symmetry, thereby enabling the numerical simulation
of dynamics induced by non-perturbative modifications to the SU($n$)
Heisenberg model.  While the latter simulations generally accumulate
uncontrolled truncation errors, we identify criteria under which
numerical results correctly capture the qualitative dynamical behavior
of a system.

@@

\begin{figure}
  \centering
  \includegraphics[valign=m]{sketch_spins.pdf}
  $\stackrel{\displaystyle\Pi_{ij}}
  {\displaystyle\longleftrightarrow}$
  \includegraphics[valign=m]{sketch_spins_swap.pdf}
  \caption{The permutation operator $\Pi_{ij}$ swaps the states of
    spins $i$ and $j$, which in the case of SU(2) can be represented
    by magnetization vectors in 3-dimensional space.  In the case of
    SU($n$), these vectors live in a higher-dimensional space without
    a clear geometric interpretation, but the interpretation of the
    permutation operator $\Pi_{ij}$ remains the same.}
  \label{fig:sketch_spins}
\end{figure}

@@

As a concrete example, this sort of scenario can be used in the SU(2)
case to generate spin squeezed states that harness entanglement to
surpass classical limits on measurement
precision\cite{he2019engineering}\red{[cite XXZ squeezing paper]}.
Starting with a general (gapped) SU(2)-symmetric Hamiltonian of the
form in \eqref{eq:H_SU2} we can consider the addition of Ising
interactions
\begin{align}
  H_{\t{Ising}} = \sum_{i\ne j} J_{ij} s_\z^{(i)} s_\z^{(j)}.
  \label{eq:Ising}
\end{align}
If these interactions are sufficiently weak for perturbative
treatment, as in the case of the XXZ model near the Heisenberg
critical point, then the effective Hamiltonian induced by
$H_{\t{Ising}}$ on initially PS states $\ket\psi\in\M_0$ is the
one-axis twisting (OAT) Hamiltonian\cite{ma2011quantum}
\begin{align}
  H_{\t{OAT}} = \chi S_\z^2,
  \label{eq:OAT}
\end{align}
where the collective operator $S_\z\equiv\sum_{j=1}^Ns_\z^{(j)}$, and
the squeezing strength $\chi$ is equal to the mean Ising coupling
strength $J_{ij}$.  If Ising interactions are unavailable, the same
effect can be achieved by applying an inhomogeneous magnetic field,
\begin{align}
  H_{\t{B}} = \sum_i B_i s_\z^{(i)},
  \label{eq:magnetic}
\end{align}
which gives rise to an OAT Hamiltonian at second order in perturbation
theory, with a squeezing strength $\chi$ that is proportional to the
variance of the magnetic field strength $B_i$.

In order to go beyond first-order treatment of two-body interactions
such as \eqref{eq:Ising}, or beyond second order in the treatment of
single-body terms such as \eqref{eq:magnetic}, it will turn out to be
necessary to introduce $M$-body operators with $M>2$.  Generalizing
the above examples to the case of SU($n$) systems, we therefore
consider a general $M$-body operator of the form
\begin{align}
  \O\p{\m w,X} \equiv \sum_{k\in\D_N\p{M}} w_k X_k,
  \label{eq:multi_body_op}
\end{align}
where $\m w$ is a dimension-$M$ (i.e.~$M$-index) tensor of scalar
coefficients $w_k\equiv w_{k_1k_2\cdots k_M}$; $X$ is an $M$-spin
operator, e.g.~$s_\z\otimes s_\z$ in the case of Ising interactions
with $M=2$; $k\equiv\p{k_1,k_2,\cdots,k_M}$ is a list of the
individual spins $k_i\in\ZZ_N$ that the operator
$X_k\equiv X_{k_1k_2\cdots k_M}$ acts on; and
\begin{align}
  \D_N\p{M} \equiv
  \set{ k \in \ZZ_N^M : \t{all entries $k_i$ of $k$ are distinct } },
  \label{eq:off_diags}
\end{align}
is the strictly ``off-diagonal'' part of $\ZZ_N^M$, which is necessary
to identify for a consistent definition of $X_k$ as an $M$-body
operator.  In this notation, we can write the magnetic field and
Hamiltonians in \eqref{eq:magnetic} and \eqref{eq:Ising} as
$H_{\t{B}}=\O\p{\m B,s_\z}$ and
$H_{\t{Ising}}=\O\p{\m J,s_\z\otimes s_\z}$.

If the addition $\O\p{\m w,X}$ to the SU($n$)-symmetric Hamiltonian
$H_0$ in \eqref{eq:H_0} is sufficiently small, then we can treat the
effect of $\O\p{\m w,X}$ on the PS manifold $\M_0$ perturbatively.
The effective Hamiltonians $H_{\t{eff}}^{(1)}$ and $H_{\t{eff}}^{(2)}$
induced by $\O\p{\m w,X}$ on $\M_0$ at first and second order in
perturbation theory are\cite{bravyi2011schrieffer}
\begin{align}
  H_{\t{eff}}^{(1)} = \P_0 \O\p{\m w,X} \P_0,
  &&
  H_{\t{eff}}^{(2)} = - \sum_{\Delta\ne0}
  \f1\Delta \P_0 \O\p{\m w,X} \P_\Delta \O\p{\m w,X} \P_0,
  \label{eq:H_eff}
\end{align}
where $\P_\Delta$ is a projector onto the eigenspace $\E_\Delta$ of
the interaction Hamiltonian $H_0$ with interaction energy $\Delta$
above that of PS manifold $\M_0$.  In particular, $\P_0$ is a
projector onto $\M_0$ itself, and $\E_0=\M_0$.  The first order
effective Hamiltonian $H_{\t{eff}}^{(1)}$ simply projects
$\O\p{\m w,X}$ onto the PS manifold $\M_0$, and takes the form
\begin{align}
  H_{\t{eff}}^{(1)} \EQPS \mean w \, \col X,
  \label{eq:H_eff_1}
\end{align}
where $\EQPS$ denotes equality under a restriction to $\M_0$; the
coefficient $\mean w$ is the average of all $w_k$; and $\col X$ is a
collective version of $X$:
\begin{align}
  \mean w \equiv \f1{\abs{\D_N\p{M}}}
  \sum_{k\in\D_N\p{M}} w_k,
  &&
  \col X \equiv \sum_{k\in\D_N\p{M}} X_k,
\end{align}
with $\abs{\D_N\p{M}}=\prod_{j=0}^{M-1}\p{N-j}$.  In the case of a
magnetic field $s_\z$ or Ising interactions $s_\z\otimes s_\z$, for
example,
\begin{align}
  \col{s_\z} = \sum_i s_\z^{(i)} = S_\z,
  &&
  \col{s_\z\otimes s_\z}
  = \sum_{i\ne j} s_\z^{(i)} s_\z^{(j)}
  = S_\z^2 - \f14 N.
\end{align}
The second order effective Hamiltonian $H_{\t{eff}}^{(2)}$ in
\eqref{eq:H_eff} takes more work to simplify due to the presence of a
projector $\P_\Delta$ onto the manifold $\E_\Delta$ of states with
excitation energy $\Delta$.  This projector essentially picks off the
part of $\O\p{\m w,X}$ that is strictly off-diagonal with respect to
the ground- and excited-state manifolds $\E_0$ and $\E_\Delta$.  We
therefore need to decompose $\O\p{\m w,X}$ into components that
generate states of definite excitation energy when acting on PS states
$\ket\psi\in\M_0$.  The SU($n$) symmetry of $H_0$ enables such a
decomposition of the form
\begin{align}
  H_0 \O\p{\m w,X} \ket\psi
  = \sum_\Delta \p{E_0+\Delta} \O\p{\m w_\Delta,X} \ket\psi,
\end{align}
where, thinking of the tensor $\m w$ as a $\abs{\D_N\p{M}}$-component
vector, the tensor $\m w_\Delta$ can be found found by diagonalizing a
matrix $\hat{\m h}$ of dimensions
$\abs{\D_N\p{M}}\times\abs{\D_N\p{M}}\sim N^M\times N^M$, and
projecting $\m w$ onto the eigenspace of $\hat{\m h}$ with eigenvalue
$\Delta$ (see Appendix \ref{sec:eigenstates}).  The matrix
$\hat{\m h}$ is determined entirely by the couplings couplings
$h_{ij}$ of $H_0$.  While the computational cost of diagonalizing
$\hat{\m h}$ is nominally $O\p{N^{3M}}$, spatial symmetries of $H_0$
and $\O\p{\m w,X}$ such as translational invariance or isotropy can
drastically reduce the complexity of this eigenvalue problem, as we
discuss in Appendix \ref{sec:symmetries}.

Equipped with the decomposition of $\O\p{\m w,X}$ into terms
$\O\p{\m w_\Delta,X}$ that generate states of definite excitation
energy $\Delta$, we can expand
\begin{align}
  H_{\t{eff}}^{(2)} = -\sum_{\Delta\ne0} \f1\Delta
  \P_0 \O\p{\m w_\Delta,X}^2 \P_0.
\end{align}
Simplifying the product $\P_0 \O\p{\m w_\Delta,X}^2 \P_0$ generally
yields $r$-body operators for all positive integers $r\le2M$.  We
provide a general prescription for simplifying such products in
Appendix \ref{sec:operator_product}, and explicitly work out the
single-body ($M=1$) and two-body ($M=2$) cases in Appendices
\ref{sec:single_pair_prod} and \ref{sec:two_pair_prod}.  If $X$ is a
single-body operator, then
\begin{align}
  H_{\t{eff}}^{(2)}
  = \sum_{\Delta\ne0} \f{\var\p{\m w_\Delta}}{\p{N-1}\Delta}
  \p{\col{X}^2 - N \col{X^2}},
  \label{eq:H_eff_2_1}
\end{align}
where
\begin{align}
  \var\p{\m w} \equiv \f1{\abs{\D_N\p{M}}}
  \sum_{k\in\D_N\p{M}} \p{w_k - \mean w}^2
\end{align}
is the variance of $\m w$.  In the SU(2) case, this result implies
that an inhomogeneous magnetic field ($X=s_\z$) yields the OAT
Hamiltonian $H_{\t{OAT}}$ in \eqref{eq:OAT} at second order
perturbation theory.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Existence of a spectral gap with long-range power-law
  interactions}
\label{sec:spectral_gap}
% todo: make a sub subsection of B.1

@@

For simplicity, we first assume that the $M$-body operator
$\O\p{\m w,X}$ obeys the same spatial symmetries as $H_0$.  For any
equivalence class $s\in\S_M$, we then define a uniform superposition
$\ket{s}$ over all $k\in s$, as well as a projector $\P_\S$ onto the
space of uniform superpositions:
\begin{align}
  \ket{s} \equiv \f1{\sqrt{\abs{s}}} \sum_{k\in s} \ket{k},
  &&
  \P_{\S_M} \equiv \sum_{s\in\S_M} \op{s},
\end{align}
where $\abs{s}$ is the number of elements in $s$.  In turn, we expand
a vector of $M$-body operator coefficients
$\m w=\sum_{k\in\D_N\p{M}}w_k\ket{k}$ in this symmetrized basis:
\begin{align}
  \m w = \P_{\S_M} \m w
  = \sum_{s\in\S_M} \sum_{k\in s} w_k \ket{s} \bk{s|k}
  = \sum_{s\in\S_M} w_s \sqrt{\abs{s}} \ket{s},
\end{align}
where all $w_k$ with $k\in s$ are equal by assumption, so we can
define $w_s\equiv w_k$ for any $k\in s$.  Finally, we expand the
projection of $\hat{\m h}$ onto the symmetric subspace:
\begin{align}
  \P_{\S_M} \hat{\m h} \P_{\S_M}
  = \sum_{s\in\S_M} \f{\ket{s}}{\sqrt{\abs{s}}} \sp{
    \sum_{\substack{k\in s\\a\in\ZZ_M}}
    \sum_{\substack{p\in\ZZ_N\\p\notin k}}
    h_{p k_a} \f{\bra{\sp{k_{a:p}}}}{\sqrt{\abs{\sp{k_{a:p}}}}}
    + \sum_{\set{a,b}\in\C_M\p{2}} h_{k_ak_b}
    \f{\bra{\sp{k_{a\lra b}}}}{\sqrt{\abs{\sp{k_{a\lra b}}}}}
    - h^{\t{D}}_s \f{\bra{s}}{\sqrt{\abs{s}}}},
\end{align}
where $h^{\t{D}}_s\equiv h^{\t{D}}_k$ for any $k\in s$, with
$h^{\t{D}}_k$ defined in \eqref{eq:multi_body_op_diag_element}; and
$\sp{k}$ denotes the equivalence class of $k$.  The reduced multi-body
eigenvalue problem is then
\begin{align}
  \P_{\S_M} \hat{\m h} \P_{\S_M} \c \m w = \Delta \m w.
\end{align}
More generally, even if the couplings $\m w$ of $\O\p{\m w,X}$ do not
obey the spatial symmetries as $H_0$, these symmetries generally endow
$\hat{\m h}$ with a block-diagonal structure that reduces the
complexity of the multi-body eigenvalue problem.  We leave this
reduction to future work.  % todo: perform reduction

@@

As a final point, we note that the choice to sum (average) for the
scalar (operator) content of $\Q_g\p{\m w,\m X}$ in
\eqref{eq:sum_assignments} and \eqref{eq:sum_diagrams} was arbitrary,
and we could just as well have chosen to perform an average (sum) for
the scalar (operator) content.  For later convenience, we therefore
define the {\it mixed} sums and averages
\begin{align}
  \ut{\m w}\p{g} \equiv \f1{\abs{\L\p{g}}}
  \sum_{\ell\in\L\p{g}} \col{\m w}\p{\ell}
  = \f1{\abs{\L\p{g}}} \sum_{\ell\in\L\p{g}}
  \sum_{k\in\D_N\p{\abs{g}}} \m w\p{k,\ell},
\end{align}
\begin{align}
  \ot{\m X}\p{g} \equiv \sum_{\ell\in\L\p{g}} \mean{\m X}\p{\ell}
  \EQPS \sum_{\ell\in\L\p{g}} \m X\p{k,\ell}
  ~\t{for any}~ k\in\D_N\p{\abs{g}}.
\end{align}
If all tensors $w_\alpha$ in $\m w$ (or all operators $X_\alpha$ in
$\m X$) are permutationally symmetric, then
\begin{align}
  \Q_g\p{\m w,\m X} \stackrel{\t{sym}}{=}_{\t{PS}}
  \ut{\m w}\p{g} \ot{\m X}\p{g}.
  \label{eq:sum_diagrams_alt}
\end{align}

>>>>

\DeclareMathOperator{\diag}{diag}

@@

In this work, we generalize some of the techniques used to harness
SU(2)-symmetric spin interactions for metrological advantage to the
case of SU($n$).  In doing so, we develop a low-energy effective
theory of gapped SU($n$) ferromagnets with weakly broken SU($n$)
symmetry, e.g.~by the presence of external fields or additional
interaction terms.  The restriction of our theory to ferromagnetic
interactions can be relaxed to the assumption of
\begin{enumerate*}
\item an initial state in the manifold of permutationally-symmetric
  states, a multi-level generalization of the Dicke manifold for
  two-level spins, together with the assumption that
\item this manifold is gapped away from states that break
  permutational symmetry.
\end{enumerate*}
In addition to working out the effective Hamiltonian induced by weak
SU($n$) symmetry breaking through second order in perturbation theory,
we develop a diagrammatic method to systematically compute matrix
elements of arbitrary spin operators within manifolds of weakly broken
permutational symmetry, thereby enabling the numerical simulation of
dynamics induced by non-perturbative modifications to the SU($n$)
Heisenberg model.  While the latter simulations generally accumulate
uncontrolled truncation errors, we identify criteria under which
numerical results correctly capture the qualitative dynamical behavior
of a system.

@@

\begin{figure}
  \centering
  \includegraphics[valign=m]{sketch_spins.pdf}
  $\stackrel{\displaystyle\Pi_{ij}}
  {\displaystyle\longleftrightarrow}$
  \includegraphics[valign=m]{sketch_spins_swap.pdf}
  \caption{The permutation operator $\Pi_{ij}$ swaps the states of
    interacting spins $i$ and $j$, which in the case of SU(2) can be
    represented by magnetization vectors in 3-dimensional space.  In
    the case of SU($n$), these vectors live in a higher-dimensional
    space without a clear geometric interpretation, but the
    interpretation of the permutation operator $\Pi_{ij}$ remains the
    same.}
  \label{fig:sketch_spins}
\end{figure}

@@

As a concrete example, this sort of scenario can be used in the SU(2)
case to generate spin squeezed states that harness entanglement to
surpass classical limits on measurement
precision\cite{he2019engineering}\red{[cite XXZ squeezing paper]}.
Starting with a general (gapped) SU(2)-symmetric Hamiltonian of the
form in \eqref{eq:H_SU2} we can consider the addition of Ising
interactions
\begin{align}
  H_{\t{Ising}} = \sum_{i\ne j} J_{ij} s_\z^{(i)} s_\z^{(j)}.
  \label{eq:Ising}
\end{align}
If these interactions are sufficiently weak for perturbative
treatment, as in the case of the XXZ model near the Heisenberg
critical point, then the effective Hamiltonian induced by
$H_{\t{Ising}}$ on initially PS states $\ket\psi\in\M_0$ is the
one-axis twisting (OAT) Hamiltonian\cite{ma2011quantum}
\begin{align}
  H_{\t{OAT}} = \chi S_\z^2,
  \label{eq:OAT}
\end{align}
where the collective operator $S_\z\equiv\sum_{j=1}^Ns_\z^{(j)}$, and
the squeezing strength $\chi$ is equal to the mean Ising coupling
strength $J_{ij}$.  If Ising interactions are unavailable, the same
effect can be achieved by applying an inhomogeneous magnetic field,
\begin{align}
  H_{\t{B}} = \sum_i B_i s_\z^{(i)},
  \label{eq:magnetic}
\end{align}
which gives rise to an OAT Hamiltonian at second order in perturbation
theory, with a squeezing strength $\chi$ that is related to the
variance of the magnetic field strength $B_i$.

Similarly to single- and two-body operators such as $H_{\t{B}}$ and
$H_{\t{Ising}}$ in \eqref{eq:magnetic} and \eqref{eq:Ising}, it will
be necessary for our purposes to additionally consider the
introduction of $M$-body operators with $M>2$ \red{[todo: explain why
  better]}. Generalizing the above examples to the case of SU($n$)
systems, we therefore consider a general $M$-body operator of the form
\begin{align}
  \O\p{\m w,X} \equiv \sum_{k\in\D_N\p{M}} w_k X_k,
  \label{eq:multi_body_op}
\end{align}
where $\m w$ is a dimension-$M$ (i.e.~$M$-index) tensor of scalar
coefficients $w_k\equiv w_{k_1k_2\cdots k_M}$; $X$ is an $M$-spin
operator, e.g.~$s_\z\otimes s_\z$ in the case of Ising interactions
with $M=2$; $k\equiv\p{k_1,k_2,\cdots,k_M}$ is a list of the
individual spins $k_i\in\ZZ_N$ that the operator
$X_k\equiv X_{k_1k_2\cdots k_M}$ acts on; and
\begin{align}
  \D_N\p{M} \equiv
  \set{ k \in \ZZ_N^M : \t{all entries $k_i$ of $k$ are distinct } },
  \label{eq:off_diags}
\end{align}
is the strictly ``off-diagonal'' part of $\ZZ_N^M$, which is necessary
to identify for a consistent definition of $X_k$ as an $M$-body
operator.  In this notation, we can write the magnetic field and Ising
Hamiltonians in \eqref{eq:magnetic} and \eqref{eq:Ising} as
$H_{\t{B}}=\O\p{\m B,s_\z}$ and
$H_{\t{Ising}}=\O\p{\m J,s_\z\otimes s_\z}$.

If the addition $\O\p{\m w,X}$ to the SU($n$)-symmetric Hamiltonian
$H_0$ in \eqref{eq:H_0} is sufficiently small, namely with operator
norm $\norm{\O\p{\m w,X}}$ less than half the spectral gap
$\Delta_{\t{gap}}$ of $H_0$, $\norm{\O\p{\m w,X}}<\Delta_{\t{gap}}/2$,
then we can treat the effect of $\O\p{\m w,X}$ on the ground-state PS
manifold $\M_0$ perturbatively.  The effective Hamiltonians
$H_{\t{eff}}^{(1)}$ and $H_{\t{eff}}^{(2)}$ induced by $\O\p{\m w,X}$
on $\M_0$ at first and second order in perturbation theory
are\cite{bravyi2011schrieffer}
\begin{align}
  H_{\t{eff}}^{(1)} = \P_0 \O\p{\m w,X} \P_0,
  &&
  H_{\t{eff}}^{(2)} = - \sum_{\Delta\ne0}
  \f1\Delta \P_0 \O\p{\m w,X} \P_\Delta \O\p{\m w,X} \P_0,
  \label{eq:H_eff}
\end{align}
where $\P_\Delta$ is a projector onto the eigenspace $\E_\Delta$ of
the interaction Hamiltonian $H_0$ with interaction energy $\Delta$
above that of PS manifold $\M_0$.  In particular, $\P_0$ is a
projector onto $\M_0$ itself, and $\E_0=\M_0$.  The first order
effective Hamiltonian $H_{\t{eff}}^{(1)}$ simply projects
$\O\p{\m w,X}$ onto the PS manifold $\M_0$, and takes the form
\begin{align}
  H_{\t{eff}}^{(1)} \EQPS \mean w \, \col X,
  \label{eq:H_eff_1}
\end{align}
where $\EQPS$ denotes equality under a restriction to $\M_0$; the
coefficient $\mean w$ is the average of all $w_k$; and $\col X$ is a
collective version of $X$:
\begin{align}
  \mean w \equiv \f1{\abs{\D_N\p{M}}}
  \sum_{k\in\D_N\p{M}} w_k,
  &&
  \col X \equiv \sum_{k\in\D_N\p{M}} X_k,
\end{align}
with $\abs{\D_N\p{M}}=\prod_{j=0}^{M-1}\p{N-j}$.  In the case of a
magnetic field $s_\z$ or Ising interactions $s_\z\otimes s_\z$, for
example,
\begin{align}
  \col{s_\z} = \sum_i s_\z^{(i)} = S_\z,
  &&
  \col{s_\z\otimes s_\z}
  = \sum_{i\ne j} s_\z^{(i)} s_\z^{(j)}
  = S_\z^2 - \f14 N.
\end{align}
The second order effective Hamiltonian $H_{\t{eff}}^{(2)}$ in
\eqref{eq:H_eff} takes more work to simplify due to the presence of a
projector $\P_\Delta$ onto the manifold $\E_\Delta$ of states with
excitation energy $\Delta$.  This projector essentially picks off the
part of $\O\p{\m w,X}$ that is strictly off-diagonal with respect to
the ground- and excited-state manifolds $\E_0$ and $\E_\Delta$.  We
therefore need to decompose $\O\p{\m w,X}$ into components that
generate states of definite excitation energy when acting on PS states
$\ket\psi\in\M_0$.  The SU($n$) symmetry of $H_0$ enables such a
decomposition of the form
\begin{align}
  H_0 \O\p{\m w,X} \ket\psi
  = \sum_\Delta \p{E_0+\Delta} \O\p{\m w_\Delta,X} \ket\psi,
\end{align}
where, thinking of the tensor $\m w$ as a $\abs{\D_N\p{M}}$-component
vector, the tensor $\m w_\Delta$ can be found by diagonalizing a
matrix $\hat{\m h}$ of dimensions
$\abs{\D_N\p{M}}\times\abs{\D_N\p{M}}\sim N^M\times N^M$, and
projecting $\m w$ onto the eigenspace of $\hat{\m h}$ with eigenvalue
$\Delta$ (see Appendix \ref{sec:eigenstates}).  The matrix
$\hat{\m h}$ is determined entirely by the couplings couplings
$h_{ij}$ of $H_0$.  While the computational cost of diagonalizing
$\hat{\m h}$ is nominally $O\p{N^{3M}}$, spatial symmetries of $H_0$
and $\O\p{\m w,X}$ such as translational invariance or isotropy can
drastically reduce the complexity of this eigenvalue problem, as we
discuss in Appendix \ref{sec:symmetries}.

Equipped with the decomposition of $\O\p{\m w,X}$ into terms
$\O\p{\m w_\Delta,X}$ that generate states of definite excitation
energy $\Delta$, we can expand
\begin{align}
  H_{\t{eff}}^{(2)} = -\sum_{\Delta\ne0} \f1\Delta
  \P_0 \O\p{\m w_\Delta,X}^2 \P_0.
\end{align}
Simplifying the product $\P_0 \O\p{\m w_\Delta,X}^2 \P_0$ generally
yields $r$-body operators for all positive integers $r\le2M$.  We
provide a general prescription for simplifying such products in
Appendix \ref{sec:operator_product}, and explicitly work out the
single-body ($M=1$) and two-body ($M=2$) cases in Appendices
\ref{sec:single_pair_prod} and \ref{sec:two_pair_prod}.  If $X$ is a
single-body operator, then
\begin{align}
  H_{\t{eff}}^{(2)}
  = \sum_{\Delta\ne0} \f{\m w_\Delta\c\m w_\Delta}{N\p{N-1}\Delta}
  \p{\col{X}^2 - N \col{X^2}}.
  \label{eq:H_eff_2_1}
\end{align}
In the SU(2) case, this result implies that an inhomogeneous magnetic
field ($X=s_\z$) yields the OAT Hamiltonian $H_{\t{OAT}}$ in
\eqref{eq:OAT} at second order perturbation theory.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Existence of a spectral gap with long-range power-law
  interactions}
\label{sec:spectral_gap}

\red{[todo: make this a subsection of Appendix
  \ref{sec:single_body_eigenstates}]}

@@

For simplicity, we first assume that the $M$-body operator
$\O\p{\m w,X}$ obeys the same spatial symmetries as $H_0$.  For any
equivalence class $s\in\S_M$, we then define a uniform superposition
$\ket{s}$ over all $k\in s$, as well as a projector $\P_\S$ onto the
space of uniform superpositions:
\begin{align}
  \ket{s} \equiv \f1{\sqrt{\abs{s}}} \sum_{k\in s} \ket{k},
  &&
  \P_{\S_M} \equiv \sum_{s\in\S_M} \op{s},
\end{align}
where $\abs{s}$ is the number of elements in $s$.  In turn, we expand
a vector of $M$-body operator coefficients
$\m w=\sum_{k\in\D_N\p{M}}w_k\ket{k}$ in this symmetrized basis:
\begin{align}
  \m w = \P_{\S_M} \m w
  = \sum_{s\in\S_M} \sum_{k\in s} w_k \ket{s} \bk{s|k}
  = \sum_{s\in\S_M} w_s \sqrt{\abs{s}} \ket{s},
\end{align}
where all $w_k$ with $k\in s$ are equal by assumption, so we can
define $w_s\equiv w_k$ for any $k\in s$.  Finally, we expand the
projection of $\hat{\m h}$ onto the symmetric subspace:
\begin{align}
  \P_{\S_M} \hat{\m h} \P_{\S_M}
  = \sum_{s\in\S_M} \f{\ket{s}}{\sqrt{\abs{s}}} \sp{
    \sum_{\substack{k\in s\\a\in\ZZ_M}}
    \sum_{\substack{p\in\ZZ_N\\p\notin k}}
    h_{p k_a} \f{\bra{\sp{k_{a:p}}}}{\sqrt{\abs{\sp{k_{a:p}}}}}
    + \sum_{\set{a,b}\in\C_M\p{2}} h_{k_ak_b}
    \f{\bra{\sp{k_{a\lra b}}}}{\sqrt{\abs{\sp{k_{a\lra b}}}}}
    - h^{\t{D}}_s \f{\bra{s}}{\sqrt{\abs{s}}}},
\end{align}
where $h^{\t{D}}_s\equiv h^{\t{D}}_k$ for any $k\in s$, with
$h^{\t{D}}_k$ defined in \eqref{eq:multi_body_op_diag_element}; and
$\sp{k}$ denotes the equivalence class of $k$.  The reduced multi-body
eigenvalue problem is then
\begin{align}
  \P_{\S_M} \hat{\m h} \P_{\S_M} \c \m w = \Delta \m w.
\end{align}
More generally, even if the couplings $\m w$ of $\O\p{\m w,X}$ do not
obey the spatial symmetries as $H_0$, these symmetries still endow
$\hat{\m h}$ with a block-diagonal structure that reduces the
complexity of the multi-body eigenvalue problem.  We leave this
reduction to future work.  % todo: perform reduction

@@

As a final point, we note that the choice to sum (average) for the
scalar (operator) content of $\Q_g\p{\m w,\m X}$ in
\eqref{eq:sum_assignments} and \eqref{eq:sum_diagrams} was arbitrary,
and we could just as well have chosen to interchange sums and
averages.  To simplify the diagrammatic formalism we develop in the
following section, we therefore define the {\it mixed} sums and
averages
\begin{align}
  \ut{\m w}\p{g} \equiv \f1{\abs{\L\p{g}}}
  \sum_{\ell\in\L\p{g}} \col{\m w}\p{\ell}
  = \f1{\abs{\L\p{g}}} \sum_{\ell\in\L\p{g}}
  \sum_{k\in\D_N\p{\abs{g}}} \m w\p{k,\ell},
\end{align}
\begin{align}
  \ot{\m X}\p{g} \equiv \sum_{\ell\in\L\p{g}} \mean{\m X}\p{\ell}
  \EQPS \sum_{\ell\in\L\p{g}} \m X\p{k,\ell}
  ~\t{for any}~ k\in\D_N\p{\abs{g}},
\end{align}
where $\ut{\m w}\p{g}$ sums over index values $k\in\D_N\p{\abs{g}}$,
but averages over index assignments $\ell\in\L\p{g}$; and vice versa
for $\ot{\m X}\p{g}$.  If all tensors $w_\alpha$ in $\m w$ (or all
operators $X_\alpha$ in $\m X$) are permutationally symmetric, then
\begin{align}
  \Q_g\p{\m w,\m X} \stackrel{\t{sym}}{=}_{\t{PS}}
  \ut{\m w}\p{g} \ot{\m X}\p{g}.
  \label{eq:sum_diagrams_alt}
\end{align}
